---
output:
  html_document:
    theme: spacelab
    mathjax: default
    code_folding: hide
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
editor_options: 
  chunk_output_type: console
params:
  echo: yes
  relative: FALSE
title: "Differential abundance analysis Melanoma"
author: "Petra Schwalie and Juliane Siebourg"
date: '`r format(Sys.time(), "%d %B, %Y")`'
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = T, echo = params$echo, results = 'asis', 
                      message = F, warning = F, error = T,
                      width = 120)
# dev = 'pdf',fig.width=12, fig.height=8, out.width='1200'  

env <- Sys.info()["sysname"]
```


```{r settings, include=FALSE}
require(RColorBrewer)
require(knitr)
require(ggthemes)
require(viridis)

library(SingleCellExperiment)
library(lme4)
library(MAST)
library(DT)

library(tidyverse)

# change ggplot theme
theme_set(theme_bw())

# To store analyis results
dir.create('results')

detach("package:dplyr", character.only = TRUE)
library("dplyr", character.only = TRUE)

```

# Load the Data

TIL data is loaded. Comparison at celltype3 and celltype2 level, NR (adjuvant only) vs. R (TF only); 
change selection for NR (nadj) vs. R for non-adjuvant analysis. Cell numbers are used as weights. 
Accounting only for gender.  Patients 38 and 12 were removed from the analysis. 

Overview of cell numbers per patient and cell type (lev1):

```{r}
mymat <- readr::read_tsv('../analyzed/sw_besca2_immune_fullMeta_res2.tsv')

mymat=mymat[,c(1:40,179,145:208)]
colnames(mymat)=make.names(colnames(mymat))
mymat=mymat[,which(!duplicated(colnames(mymat)))]

mymat <- mymat %>%
  mutate(cell_names=celltype3_pub)  %>%
  mutate(sscell_group=celltype2_pub) %>%
  mutate(cell_group=celltype1) %>%
  mutate(Lesion = replace(Lesion, Lesion == "Brain", "Other")) %>%
  mutate(Lesion = replace(Lesion, Lesion == "Subc", "Other")) %>%
  mutate(Lesion = replace(Lesion, Lesion == "Sinon", "Other")) %>%
  mutate(Lesion = replace(Lesion, Lesion == "Lung", "Other")) 
    
mymat=mymat[mymat$RCat%in%c("R","NR_nadj"),]

mymat_summary <- mymat %>% select(PatientID, experiment, Response, Respond,Lesion,Gender,Age,cell_names, sscell_group, cell_group, Sample.type, score_ProlifMyeloid_scanpy) %>% 
  group_by(PatientID, Gender,Age,Lesion, cell_names, sscell_group,Respond, Sample.type, cell_group, Response) %>% 
  summarize(mean = mean(score_ProlifMyeloid_scanpy, na.rm = TRUE), 
            median = median(score_ProlifMyeloid_scanpy, na.rm = TRUE), 
            sd = sd(score_ProlifMyeloid_scanpy, na.rm = TRUE), 
            iqr = IQR(score_ProlifMyeloid_scanpy, na.rm = TRUE),
            count = n()) %>% 
  group_by(PatientID,Respond) %>% 
  mutate(total_count = sum(count),
         freq = count/total_count) %>% 
  ungroup() %>% 
  mutate(sample_id = interaction(PatientID, cell_names)) %>% 
  mutate(sample_type = factor(Respond, levels = c('PD', 'R')))

mymat_summary <- mymat_summary %>%
  mutate(PatientID=as.character(PatientID))

chosenid=colnames(mymat_summary)[which(!colnames(mymat_summary)%in%c("mean","median","sd","iqr","MyTcFreq","total_MyTc","cDCMyFreq"      ,"total_cDCMy", "TcEMvsEx","total_TcEMvsEx","cDC12Freq","total_cDC12"))]
datatable(mymat_summary[,chosenid])

```

Overview of cell numbers per patient and cell type (lev2):

```{r}
# summary of only high level celltypes
#  dplyr::filter(celltype != 'AI10 dominated mixture') %>% 
mymat_hl_summary <- mymat %>% 
  select(PatientID, sscell_group, cell_group, experiment, Response, Respond,Lesion,Gender,Age, score_ProlifMyeloid_scanpy) %>% 
  group_by(PatientID, sscell_group, cell_group,Lesion,Gender,Age,Respond, Response) %>% 
  summarize(mean = mean(score_ProlifMyeloid_scanpy, na.rm = TRUE), 
            median = median(score_ProlifMyeloid_scanpy, na.rm = TRUE), 
            sd = sd(score_ProlifMyeloid_scanpy, na.rm = TRUE), 
            iqr = IQR(score_ProlifMyeloid_scanpy, na.rm = TRUE),
            count = n()) %>% 
  group_by(PatientID, Respond) %>% 
  mutate(total_count = sum(count),
         freq = count/total_count) %>% 
  ungroup() %>% 
  mutate(sample_id = interaction(PatientID, sscell_group)) %>% 
  mutate(sample_type = factor(Respond, levels = c('PD', 'R')))

mymat_hl_summary <- mymat_hl_summary %>%
  mutate(PatientID=as.character(PatientID))

col_tissue <- c(PD = 'cornflowerblue', R = 'orange')

chosenid=colnames(mymat_hl_summary)[which(!colnames(mymat_hl_summary)%in%c("mean","median","sd","iqr","MyTcFreq","total_MyTc","cDCMyFreq" ,"total_cDCMy", "TcEMvsEx","total_TcEMvsEx","cDC12Freq","total_cDC12"))]
datatable(mymat_hl_summary[,chosenid])

```


Available cell types:

```{r}
mymat_summary %>% select(cell_names, sscell_group, cell_group) %>% 
  unique() %>% datatable()

```

# Population counts and fractions

## Visualization per celltype and tissue 

### Total cell counts {.tabset .tabset-pills}

```{r, fig.width=4, fig.height=4}
ggplot(mymat_summary, aes(x = sample_type, y = total_count)) +
  geom_boxplot(color = 'gray') +
  geom_point(aes(color = PatientID, shape = sample_type)) +
  geom_line(aes(group = PatientID), color = 'darkgray') +
  scale_color_viridis_d()
```

Responders have overall higher cell numbers. 

### Cell counts by cell type {-}

```{r, fig.width=7, fig.height=5}
ggplot(mymat_hl_summary, aes(x = sample_type, y = count, 
                         color = PatientID, shape = sample_type)) +
  geom_point() +
  geom_line(aes(group = PatientID), color = 'darkgray') +
  facet_wrap(~sscell_group, scale = 'free_y') +
  scale_color_viridis_d()
```

```{r, fig.width=9, fig.height=9}
ggplot(mymat_summary, aes(x = sample_type, y = count, 
                         color = PatientID, shape = sample_type)) +
  geom_point() +
  geom_line(aes(group = PatientID), color = 'darkgray') +
  facet_wrap(~cell_names, scale = 'free_y') +
  scale_color_viridis_d()
```

### Cell fractions by cell type {-}

```{r, fig.width=7, fig.height=5}
ggplot(mymat_hl_summary, aes(x = sample_type, y = freq, 
                         color = PatientID, shape = sample_type)) +
  geom_point() +
  geom_line(aes(group = PatientID), color = 'darkgray') +
  facet_wrap(~sscell_group, scale = 'free_y') +
  scale_color_viridis_d()
```

```{r, fig.width=9, fig.height=9}
ggplot(mymat_summary, aes(x = sample_type, y = freq, 
                         color = PatientID, shape = sample_type)) +
  geom_point() +
  geom_line(aes(group = PatientID), color = 'darkgray') +
  facet_wrap(~cell_names, scale = 'free_y') +
  scale_color_viridis_d()
```


## Modeling population sizes

### GLMM

```{r, include=FALSE}

# wrapper function from Malgorzata 2017... 
differential_abundance_wrapper <- function(counts, md, formula, K){
  ## Fit the GLMM for each cluster separately
  ntot <- colSums(counts)
  fit_binomial <- lapply(1:nrow(counts), function(i){

    data_tmp <- data.frame(y = as.numeric(counts[i, md$sample_id]),
      total = ntot[md$sample_id], md)

    fit_tmp <- glmer(formula, weights = total, family = binomial,
      data = data_tmp)

    ## Fit contrasts one by one
    out <- apply(K, 1, function(k){
      contr_tmp <- glht(fit_tmp, linfct = matrix(k, 1))
      summ_tmp <- summary(contr_tmp)
      pval <- summ_tmp$test$pvalues
      return(pval)
    })
     return(out)
  })
  pvals <- do.call(rbind, fit_binomial)
  colnames(pvals) <- paste0("pval_", contrast_names)
  rownames(pvals) <- rownames(counts)
  ## Adjust the p-values
  adjp <- apply(pvals, 2, p.adjust, method = "BH")
  colnames(adjp) <- paste0("adjp_", contrast_names)
  return(list(pvals = pvals, adjp = adjp))
}

```

#### Patient as random effect

##### all celltypes

```{r}

avail <- unique(mymat_summary %>% select(cell_names, count) %>%   
                  group_by(cell_names) %>% 
                  summarise(count = mean(count))%>% dplyr::filter(count>20)  %>% .$cell_names)

 
```



```{r}
glme_res <- mymat_summary %>% 
  dplyr::filter(cell_names %in% avail) %>% 
  group_by(cell_names) %>% nest() %>% 
  mutate(fit = map(data, ~lme4::glmer(count/total_count ~ sample_type +(1 | PatientID), # 
                        weights = total_count, # weight by number of observation to prioritize patients with high counts?
                        family = 'binomial', 
                        data = .x)),
         #summ = map(fit, ~summary() %>% broom::tidy()),
         res = map(fit, ~multcomp::glht(.x) %>% summary() %>% broom::tidy()),
         ci =  map(fit, ~multcomp::glht(.x) %>% confint() %>% broom::tidy())
  ) #

res <- glme_res$res
names(res) <- glme_res$cell_names

toexcl=res[[1]]$contrast[which(!res[[1]]$contrast%in%c("sample_typeR","sample_typeNR","sample_typeTF"))]
res <- res %>% bind_rows(.id = 'cell_names') %>% 
  dplyr::filter(!contrast %in% toexcl) 

  
ci <- glme_res$ci
names(ci) <- glme_res$cell_names
ci <- ci %>% bind_rows(.id = 'cell_names') %>% 
  dplyr::filter(!contrast %in% toexcl) %>% 
  select(-contrast, -estimate)


res <- res %>% 
  full_join(ci) %>% 
  arrange(adj.p.value) %>% 
  mutate(cell_names = factor(cell_names, levels = (res$cell_names)),
         odds_ratio = exp(estimate),
         ci_odds_low = exp(conf.low), 
         ci_odds_high = exp(conf.high))
         #prob = odds / (1 + odds)) 

res$cell_names <- factor(res$cell_names, levels = rev(res$cell_names))
 
```

Estimated difference with confidence intervals, ordered by significance.

```{r, fig.width=4, fig.height=5}

ggplot(res, aes(x = estimate, y = cell_names)) +
  geom_vline(xintercept = 0, color = 'gray', linetype = 'dashed') +
  geom_point() +
  geom_errorbarh(aes(xmax = conf.high, xmin = conf.low), height = .1) +
  ggtitle('logOdds ratio')


```

Results table 

```{r}
res %>% 
  mutate(adj.p.value = format(adj.p.value, digits = 2)) %>% 
  kable(digits = 2) %>% 
  kableExtra::kable_styling(full_width = FALSE)

```


##### higher level celltypes

```{r}

avai2 <- unique(mymat_hl_summary %>% select(sscell_group, count) %>%   
                  group_by(sscell_group) %>% 
                  summarise(count = mean(count))%>% dplyr::filter(count>20)  %>% .$sscell_group)

```



```{r}
#Lesion+
glme_res <- mymat_hl_summary %>% 
  dplyr::filter(sscell_group %in% avai2) %>% 
  group_by(sscell_group) %>% nest() %>% 
  mutate(fit = map(data, ~lme4::glmer(count/total_count ~ sample_type+(1 | PatientID), 
                        weights = total_count, # weight by number of observation to prioritize patients with high counts?
                        family = 'binomial', 
                        data = .x)),
         res = map(fit, ~multcomp::glht(.x) %>% summary() %>% broom::tidy()),
         ci =  map(fit, ~multcomp::glht(.x) %>% confint() %>% broom::tidy())
  )


res_hl <- glme_res$res
names(res_hl) <- glme_res$sscell_group

toexcl=res_hl[[1]]$contrast[which(!res_hl[[1]]$contrast%in%c("sample_typeR","sample_typeNR","sample_typeTF"))]

res_hl <- res_hl %>% bind_rows(.id = 'sscell_group') %>% 
  dplyr::filter(!contrast %in% toexcl) %>% 
  select( -contrast)

ci <- glme_res$ci
names(ci) <- glme_res$sscell_group
ci <- ci %>% bind_rows(.id = 'sscell_group') %>% 
  dplyr::filter(!contrast %in% toexcl) %>% 
  select( -contrast, -estimate)

res_hl <- res_hl %>% 
  full_join(ci) %>% 
  arrange(adj.p.value) %>% 
  mutate(highlevel_celltype = factor(sscell_group, levels = (res_hl$sscell_group)),
         odds_ratio = exp(estimate),
         ci_odds_low = exp(conf.low), 
         ci_odds_high = exp(conf.high))
         #prob = odds / (1 + odds)) 

res_hl$sscell_group <- factor(res_hl$sscell_group, levels = rev(res_hl$sscell_group))
```

Estimated difference with confidence intervals, ordered by significance.

```{r, fig.width=4, fig.height=5}
ggplot(res_hl, aes(x = estimate, y = sscell_group)) +
  geom_vline(xintercept = 0, color = 'gray', linetype = 'dashed') +
  geom_point() +
  geom_errorbarh(aes(xmax = conf.high, xmin = conf.low), height = .1) +
  ggtitle('logOdds ratio')



```

Results table 

```{r}
res_hl %>% 
  mutate(padj = format(adj.p.value, digits = 2)) %>% 
  kable(digits = 2) %>% 
  kableExtra::kable_styling(full_width = FALSE)

```


### Wilcoxon test

```{r}

wilk_res <- mymat_summary %>% 
  dplyr::filter(cell_names %in% avail) %>% 
  spread(key = 'sample_type', value = freq) %>% 
  group_by(cell_names) %>% 
  mutate(stat = wilcox.test(PD, R) %>% .$statistic,
         p.value = wilcox.test(PD, R) %>% .$p.value) %>% 
  ungroup()


res_w <- wilk_res %>% 
  select(cell_names, stat, p.value) %>% unique() %>% 
  arrange(p.value) %>% 
  mutate(padj = p.adjust(p.value, method = 'fdr')) %>% 
  mutate(celltype = factor(cell_names, levels = (res$cell_names)))

```

Results table 

```{r}
res_w %>% 
  mutate(p.value = format(p.value, digits = 2)) %>% 
  mutate(padj = format(padj, digits = 2)) %>% 
  kable(digits = 2) %>% 
  kableExtra::kable_styling(full_width = FALSE)

```

### Comparison of raw p-values

```{r, fig.width=4, fig.height=4}

res %>% select(cell_names, adj.p.value, statistic) %>% 
  full_join(res_w %>% select(cell_names, p.value, stat), by = 'cell_names') %>% 
  ggplot(aes(x = adj.p.value, y = p.value)) + geom_point() +
  xlab('glmer') + ylab('wilcoxon') #+
```

